Enabling Ambient Backscatter \\Using Low-Cost Software Defined Radio

### 2017-04-07 paper.tex
Changed title.
Added backscatter bacground and cleaned up little

In this background section we introduce the necessary knowledge about sofware defined radios and the used hardware, the \textit{RTL-SDR}.

\subsection{Software Defined Radio}
Traditionally basic radio components like mixers or filters have been realized
in hardware. Nowadays as soon as one works in the baseband frequency the
computational resources of a personal computer are sufficient to realize these
components with signal processing done in software. \cp{We need to expand on
this introduction to SDR, assuming we need it and not just assume everyone
knows what that is.}

\subsection{RTL2832U}
\label{sub:rtl2832}
 \begin{figure}[h]
 \centering
\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q demodulator
\textit{Raeltek RTL2832U} (left IC) and the tuner with integrated LNA
\textit{Rafael Micro R820T/2} (right IC).}
 \label{fig:receiver_arch}
\end{figure}
The \textit{Realtek RTL2832U} is a terrestrial digital video broadcast
(DVB-T) demodulator IC, which is the main part of a wide range of
USB-based DVB-T receiver dongles. One prominent feature of this chip is
that it allows to retrieve the raw I/Q samples via USB. This is
originally intended for the chip to work as a simple DAB/FM software
receiver. Ham radio enthusiasts have combined their efforts to create a
software driver (the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr})
that allows DVB-T dongles based on the \textit{RTL2832} to be converted
into low-cost wideband SDR receivers. The cost of traditional SDRs has
been in the range of a few hundred to thousands of USD. Albeit much less
powerful than a typical SDR, a \textit{RTL2832U}-based DVB-T receiver
can be bought with antenna for less than 10 USD.  A RTL-SDR does not
only consist of the \textit{RTL2832U}, but also a tuner. The tuner is an
essential component, as it allows the user to select the received
frequency and outputs that signal at an intermediate frequency (IF). The
IF can then be fed into the demodulator for digitalization. In the case
of the used hardware this tuner is a \textit{Rafael Micro R820T/2} which
offers a tuning range from 42 to 1002 MHz~\cite{rafael_r820t_2011}. It
subsumes a low-noise amplifier (LNA) and filters. It presents a low
noise figure of 3.5 dB and features 65 dBc of image rejection. Figure
\ref{fig:receiver_arch}, shows a photograph of the RTL-SDR used in our
work.

@@

The received signal can be interpreted as
\begin{equation}
	s_{\text{IF}}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine (and a sine as well) from a local oscillator.
\begin{multline}
        s_{\text{IF}}(t) \cdot s_\text{L}(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{\text{IF}}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting in-phase part (I) in this case is
represented by a DC value after mixing. With a low-pass filter this DC
value can be separated from the undesired rest. An analogous procedure
is done with the quadrature part (Q) when mixing with a sine.

>>>>

In this section we introduce some necessary background related to our
work.

%\subsection{Software Defined Radio}
%Traditionally basic radio components like mixers or filters have been realized
%in hardware. Nowadays as soon as one works in the baseband frequency the
%computational resources of a personal computer are sufficient to realize these
%components with signal processing done in software. \cp{We need to expand on
%this introduction to SDR, assuming we need it and not just assume everyone
%knows what that is.}

\subsection{Backscatter Transmissions}
\input{backscatter}

\subsection{RTL2832U}
\label{sub:rtl2832}

\begin{figure}[h]
				\centering
				\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
				\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q
				demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner
				with integrated LNA \textit{Rafael Micro R820T/2} (right IC).} 
				\label{fig:receiver_arch}
\end{figure}
The \textit{Realtek RTL2832U} is a terrestrial digital video broadcast
(DVB-T) demodulator IC, which is the main part of a wide range of
USB-based DVB-T receiver dongles. One prominent feature of this chip is
that it allows to retrieve the raw I/Q samples via USB. This is
originally intended for the chip to work as a simple DAB/FM software
receiver. Ham radio enthusiasts have combined their efforts to create a
software driver (the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr})
that allows DVB-T dongles based on the \textit{RTL2832} to be converted
into low-cost wideband SDR receivers. The cost of traditional SDRs has
been in the range of a few hundred to thousands of USD. Albeit much less
powerful than a typical SDR, a \textit{RTL2832U}-based DVB-T receiver
can be bought with antenna for less than 10 USD.

An RTL-SDR also contains a tuner that allows the user to select the
received frequency.  In our case the tuner is a \textit{Rafael Micro
R820T/2} which offers a tuning range from 42 to 1002
MHz~\cite{rafael_r820t_2011}. Figure \ref{fig:receiver_arch}, shows a
photograph of the RTL-SDR used in our work.

@@

The received signal can be interpreted as
\begin{equation}
	s_{\text{IF}}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine of the carrier frequency from a local oscillator.
\begin{multline}
        s_{\text{IF}}(t) \cdot s_\text{L}(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{\text{IF}}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting in-phase part (I) in this case is
represented by a DC value after mixing. With a low-pass filter this DC
value can be separated from the undesired rest. An analogous procedure
is done with the quadrature part (Q) when mixing with a sine.
